GSE74548: Gene Methylation Enrichment Analysis

### 2022-05-14 gmea.Rmd
optimise speed

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")

@@

@@

dm <- read.table("dma3a.tsv.gz")

@@

@@

# set cores to used for parallel execution
CORES= detectCores()

# make a vector of unique gene names
gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
gnl <- strsplit( dm$UCSC_RefGene_Name ,";")

# run the test (self contained)
gme_res <- mclapply( 1:length(gn), function(i) {
  g <- gn[i]
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  tstats <- dm[myrows,"t"]
  tstats <- tstats[order(tstats)]
  myn <- length(tstats)
  mymean <- mean(tstats)
  mymedian <- median(tstats)
  wtselfcont <- wilcox.test(tstats)

  res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
           "p-value(selfcontained)"=wtselfcont$p.value)
} , mc.cores=CORES)

# format the results
gme_res_df <- do.call(rbind, gme_res)
rownames(gme_res_df) <- gme_res_df[,1]
gme_res_df <- gme_res_df[,-1]
tmp <- apply(gme_res_df,2,as.numeric)
rownames(tmp) <- rownames(gme_res_df)
gme_res_df <- as.data.frame(tmp)

# results are sorted by the sum of -log10 of the p-values
gme_res_df$sig <- -log10(gme_res_df[,4])

gme_res_df <- gme_res_df[order(-gme_res_df$sig),]

gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)

gme_res_df_wholegene <- gme_res_df

write.table(gme_res_df_wholegene,file="gmea_wholegene.tsv")

@@

head(gme_res_df,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

@@

# volcano selfcont
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
plot(gme_res_df$median , -log10(gme_res_df$`p-value(selfcontained)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(selfcontained)`) ,
  pch=19, cex=0.5, col="red")

@@

par(mfrow=c(1,2))
n=50
# self contained
gs <- head(rownames(gme_res_df[order(gme_res_df$`p-value(selfcontained)`),]),n)
tstats <- mclapply(gs, function(g) {
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  dm[myrows,"t"]
},mc.cores=CORES)
names(tstats) <- gs
tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)

if ( length(gs) >2 ) {
  tstats <- mclapply(gs, function(g) {
    myrows <- which(lapply(gnl,function(x) {
      length(which(x==g))
    } ) > 0 )
    dm[myrows,"t"]
  },mc.cores=CORES)
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="biggest effect size(median)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
} else {
  plot(1)
  mtext("too few significant genes found")
}

@@

dmscore <- data.frame( gme_res_df$median * gme_res_df$sig)

rownames(dmscore) <- rownames(gme_res_df)

@@

res <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")

head(res$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>%
  kable_paper("hover", full_width = F)

mitch_report(res,outfile="dma3a_mitch.html",overwrite=TRUE)

@@

dm <- read.table("dma3a.tsv.gz")
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]

# testing
#dm <- head(dm,20000)

dm$adj.P.Val <- p.adjust(dm$P.Value)

@@

# set cores to used for parallel execution
CORES= detectCores()

# make a vector of unique gene names
gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
gnl <- strsplit( dm$UCSC_RefGene_Name ,";")

# run the test
gme_res <- mclapply( 1:length(gn), function(i) {
  g <- gn[i]
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  tstats <- dm[myrows,"t"]
  tstats <- tstats[order(tstats)]
  myn <- length(tstats)
  mymean <- mean(tstats)
  mymedian <- median(tstats)
  wtselfcont <- wilcox.test(tstats)

  res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
           "p-value(selfcontained)"=wtselfcont$p.value)
} , mc.cores=CORES)

# format the results
gme_res_df <- do.call(rbind, gme_res)
rownames(gme_res_df) <- gme_res_df[,1]
gme_res_df <- gme_res_df[,-1]
tmp <- apply(gme_res_df,2,as.numeric)
rownames(tmp) <- rownames(gme_res_df)
gme_res_df <- as.data.frame(tmp)

# results are sorted by the sum of -log10 of the p-values
gme_res_df$sig <- -log10(gme_res_df[,4])

gme_res_df <- gme_res_df[order(-gme_res_df$sig),]

gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)

gme_res_df_promoter <- gme_res_df

write.table(gme_res_df_promoter ,file="gmea_promo.tsv")

@@

res <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")

head(res$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>%
  kable_paper("hover", full_width = F)

mitch_report(res,outfile="dma3a_mitch_promo.html",overwrite=TRUE)

## Multi-mitch

```{r,dm1_mitch_promoter}

dmscore_wholegene <- data.frame( gme_res_df_wholegene$median * gme_res_df_wholegene$sig )
rownames(dmscore_wholegene) <- rownames(gme_res_df_wholegene)
colnames(dmscore_wholegene) <- "wholegene"

dmscore_promoter <- data.frame( gme_res_df_promoter$median * gme_res_df_promoter$sig )
rownames(dmscore_promoter) <- rownames(gme_res_df_promoter)
colnames(dmscore_promoter) <- "promoter"

mm <- merge(dmscore_wholegene,dmscore_promoter,by=0)
rownames(mm) <- mm[,1]
mm[,1] = NULL

mmres <- mitch_calc(x=mm, genesets=genesets,priority="effect")

head(mmres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (wholegene and promoter)") %>%
  kable_paper("hover", full_width = F)

#mitch_report(mmres,outfile="mitch_multi.html",overwrite=TRUE)

>>>>

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("IlluminaHumanMethylation450kmanifest")
library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
library("tictoc")

## Get array annotation data

```{r,anno}

ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

@@

## Read in and inspect the data

@@

dm <- read.table("dma3a.tsv.gz")
dm <- dm[,4:9]
dm <- merge(myann,dm,by=0)
rownames(dm) <- dm[,1]
dm[,1] = NULL

@@

# trim down the input dataset
dm <- dm[,c("UCSC_RefGene_Name","t")]

@@

```{r,wg}
# set cores to used for parallel execution
CORES= detectCores()

calc_sc <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a[2],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"] 
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
gme_res_wholegene <- calc_sc(dm)
time2 <- toc() #38 44 41 40 44 41 40
time2
df <- gme_res_wholegene[[1]]
res <- gme_res_wholegene[[2]]
write.table(res,file="gmea_wholegene.tsv")

@@

head(res,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

@@

# volcano selfcont
sig <- subset(res,`fdr(sc)` < 0.05)
plot(res$median , -log10(res$`p-value(sc)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(sc)`) ,
  pch=19, cex=0.5, col="red")

@@

par(mfrow=c(1,2))
n=50
# self contained

gs <- head(rownames(res),50)
tstats <- lapply(gs, function(g) {
  df[which(df$genes==g),"tvals"]
})
names(tstats) <- gs

tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(res,`fdr(sc)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)

if ( length(gs) >2 ) {
  tstats <- lapply(gs, function(g) {
    df[which(df$genes==g),"tvals"]
  })
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="biggest effect size(median)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
} else {
  plot(1)
  mtext("too few significant genes found")
}

@@

dmscore <- data.frame( res$median * res$sig)

rownames(dmscore) <- rownames(res)

@@

mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>%
  kable_paper("hover", full_width = F)

mitch_report(mres,outfile="dma3a_mitch.html",overwrite=TRUE)

@@

dm <- read.table("dma3a.tsv.gz")
dm <- dm[,4:9]
dm <- merge(myann,dm,by=0)
rownames(dm) <- dm[,1]
dm[,1] = NULL
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]

@@

```{r,promo2}

# set cores to used for parallel execution

tic()
gme_res_promoter <- calc_sc(dm)
time2 <- toc() #15.7 20.2 20.3 17.5 17.9 17.7 20.6 
time2

df <- gme_res_promoter[[1]]
res <- gme_res_promoter[[2]] 
write.table(res ,file="gmea_promo.tsv")

@@

mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>%
  kable_paper("hover", full_width = F)

mitch_report(mres,outfile="dma3a_mitch_promo.html",overwrite=TRUE)
